White Wine Exploration by Elisa Marchione
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv('wineQualityWhites.csv')
```

> This project explores a dataset that contains 4,898 white wines.  

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

> There are 11 variables that quantify the chemical properties of the wine. There is a rating for each wine between 0 (very bad) and 10 (very excellent), averaged from at least 3 wine experts.

```{r echo=FALSE}
# List all the variables
names(wine)
```

> This data set was already cleaned up, so the data types are appropriate. 

```{r echo=FALSE}
# Check the data types of all variables
str(wine)
```

> First of all, I want to see the distribution of the quality scores for all the white wines. 

```{r echo=FALSE}
# Quick histogram of quality
qplot(data = wine, x = quality)
```

```{r echo=FALSE, Univariate_plots}
# Update histogram for better binwidth
qplot(data = wine, 
      x = quality,
      binwidth = .5)
```

```{r echo=FALSE}
summary(wine$quality)
```

> It appears to have a normal distribution, with 6 being the most common score. 

> Our guiding question with this dataset is to determine which chemical properties influence the quality of white wines. I'll start out by running some summary statistics to look for any variables that stand out. 

```{r echo=FALSE}
# Summary stats for all variables
summary(wine)
```

> The variables that stand out to me at first are the ones that have a wide range between min and max, such as residual.sugar, free.sulfur.dioxide, total.sufur.dioxide. 

```{r echo=FALSE, Univariate_plots}
# Quick histogram of residual.sugar
qplot(data = wine, x = residual.sugar)
```

> I'm limiting the data to see the bulk of the information, as well as updating the bin width to 1 so I can get a better idea of the distribution. 

```{r echo=FALSE}
# Update plot to remove outliers
qplot(data = wine, 
      x = residual.sugar,
      binwidth = 1,
      xlim = c(0,25))
```

> It appears that residual.sugar is right-skewed, with most of the values being very low.

> Domain research has determined that residual.sugar actually determines the type of wine. I realized that I could create a new variable called type from the residual.sugar value and create a categorical value that might be useful for further analysis. 

```{r echo=FALSE}
# Create new variable and set categories
wine$type <- "Other"
wine$type[wine$residual.sugar<10] <- "Dry"
wine$type[10<=wine$residual.sugar & wine$residual.sugar<19] <- "Off Dry"
wine$type[19<=wine$residual.sugar & wine$residual.sugar<51] <- "Semi Sweet"
wine$type[51<=wine$residual.sugar & wine$residual.sugar<121] <- "Medium Sweet"
wine$type[121<=wine$residual.sugar] <- "Sweet"

# Convert into factor and set order
wine$type <- factor(wine$type, levels = c("Dry", "Off Dry", "Semi Sweet", "Medium Sweet", "Sweet"), ordered = TRUE)

# Quick plot of type values
qplot(data = wine, x = type)

```

> It appears that a large majority of the wines studied were considered dry and off dry. 

> I want to see how some of the other statistics are distributed. 

```{r echo=FALSE}
# Quick histogram of free.sulfur.dioxide
qplot(data = wine, x = free.sulfur.dioxide)
```

> Free.sulfur.dioxide appears to have a normal distribution. 

```{r echo=FALSE}
# Quick histogram of total.sulfur.dioxide
qplot(data = wine, x = total.sulfur.dioxide)
```

> Total.sufulr.dioxide also appears to have a normal distribution. 

```{r echo=FALSE}
# Quick histogram of pH
qplot(data = wine, x = pH)
```

> pH has a normal distribution.

```{r echo=FALSE}
# Quick histogram of fixed.acidity
qplot(data = wine, x = fixed.acidity,
      binwidth = .5)
```

> Fixed.acidity has a normal distribution.

```{r echo=FALSE}
# Quick histogram of volatile.acidity
qplot(data = wine, x = volatile.acidity,
      binwidth = .05)
```

> Volatile.acidity has a relatively normal distribution.

> Some domain research has helped me understand that Fixed acidity is a formula of total acidity minus volatile acidity. In this dataset, we are keeping track of both fixed.acidity and volatile.acidity, so we should we able to calculate a new variabled called total.acidity. 

```{r echo=FALSE}

wine$total.acidity <- wine$fixed.acidity + wine$volatile.acidity

# Quick histogram of total.acidity
qplot(data = wine, x = total.acidity)
```

> As expected, there is also a normal distribution for total.acidity. 

```{r echo=FALSE}
# Quick histogram of citric.acid
qplot(data = wine, x = citric.acid,
      binwidth = .02)
```
> Citric.acid appears to have a couple of different peaks that I might be interested in diving into more.

```{r echo=FALSE}
# Quick histogram of chlorides
qplot(data = wine, x = chlorides)
```

```{r echo=FALSE}
# Updating histogram of chlorides with log10
qplot(data = wine, x = chlorides, binwidth = 0.02) +
      scale_x_log10()
```

> After updating the plot with a log10 x value, I get a more normal distribution, but there are a couple of interesting peaks and valleys.

```{r echo=FALSE}
# Quick histogram of sulphates
qplot(data = wine, x = sulphates,
      binwidth = 0.05)
```

> Sulphates have a normal distribution. 

```{r echo=FALSE}
# Quick histogram of density
qplot(data = wine, x = density,
      binwidth = .001)
```
> Density appears to have a normal distribution as well, with a couple of outliers. 

```{r echo=FALSE}
# Quick histogram of alcohol
qplot(data = wine, x = alcohol)
```

> Updating the binwidth to get a better idea of the distribution for alcohol level. I found that using a binwidth of .1 works well since the alcohol value generally has one decimal.

```{r echo=FALSE}
# Updating plot to get better binwidth
qplot(data = wine, 
      x = alcohol,
      binwidth = .1)
```

> I'm not quite sure if this would be considered multimodal or not. There are quite a lot of peaks and valleys in the distribution of alcohol levels. 

> I want to take a look at the subsets of quality scores to get a better idea what alcohol level variety there is among them. 

```{r echo=FALSE}
# Subset of Worst Quality

worst <- subset(wine, quality < 6)
qplot(data = worst, x = alcohol,
      binwidth = .3)
```

> For the worst quality wines, the distribution of alcohol levels was more right-skewed.

```{r echo=FALSE}
# Subset of Medium Quality

med <- subset(wine, quality = 6)
qplot(data = med, x = alcohol,
      binwidth = .3)
```

> For the medium quality wines, the distribution was still right-skewed, with lower alcohol values being more abundant. 

```{r echo=FALSE}
# Subset of Best Quality

best <- subset(wine, quality > 6)
qplot(data = best, x = alcohol,
      binwidth = .3)
```

> For the best quality wines, the distribution was bi-modal, with alcohol values being higher than the worst quality wines. 

```{r echo=FALSE}
# Summary of worst quality alcohol levels
summary(worst$alcohol)
```

```{r echo=FALSE}
# Summary of medium quality alcohol levels
summary(med$alcohol)
```

```{r echo=FALSE}
# Summary of best quality alcohol levels
summary(best$alcohol)
```

> I will definitely want to keep diving into the connection between alcohol level and quality in my later analyses. 

# Univariate Analysis

### What is the structure of your dataset?

> There are 4,898 different white vines with 12 variables (fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, and quality). 
*** added variables - total.acidity, type

> Observations:
*** range in quality
*** Type order, derived from residual.sugar amount
*** Median quality = 6
*** Normal distributions for a lot of the chemical properties
*** pH is apparently a quantitative assessment of fixed acidity. *** Total acidity is formula with fixed + volatile

### What is/are the main feature(s) of interest in your dataset?
>
*** Alcohol level and maybe residual sugar. I want to see how these fit in with other chemical properties. 
*** acidity - now that I know total acidity is combination of other variables. Domain research shows that acidity is a big part of the taste. Too little acidity leads to dull, flat, boring taste. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
> *** Domain research has shown that acidity and pH work together to create flavor and texture. I want to see how these variables interact. 


### Did you create any new variables from existing variables in the dataset?
> *** Subsets of quality to help bucket the information. 
*** Total acidity
*** Wine Type

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
> *** residual sugar, alcohol


# Bivariate Plots Section

```{r}
install.packages('corrplot')
```


```{r echo=FALSE, Bivariate_Plots}
library(corrplot)
library(RColorBrewer)

W <- cor(wine, use='complete.obs')
corrplot(W, type='upper', order='hclust',
        col=brewer.pal(n=8, name='RdYlBu'))
```

> *** correlation matrix

>*** ph vs total acidity

> *** quality vs total acidity

> *** quality vs alcohol content

> *** type vs total acidity

> *** look for other interesting correlations with matrix






# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!